---
title: "How the concordance tables are built"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How the concordance tables are built}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(concordance990)
library(data.table)
```

The package generates the one-row-per-xpath concordance, in the v1 layout and
in a v2 layout, from the component tables described in
`vignette("table-structure")`. This page shows the steps and the functions
data users and parsers call.

## From components to one row per xpath

`build_concordance()` does five things:

1. **Join** each xpath to its variable (`xpaths.csv` + `variables.csv`).
   Variable-level attributes (table, label, description, location code,
   scope, data type) come from the variable.
2. **Apply the xpath overrides**: where v1 recorded a different value for one
   xpath, that value replaces the variable's (`xpath_overrides.csv`).
3. **Add the table attributes**: cardinality (`rdb_relationship`) and part.
4. **Select a form**, if asked (below).
5. **Order** the rows as in v1 (`v1_order`, new xpaths at the end) and select
   the columns: `format = "v1"` returns the v1 layout (`concordance_columns`:
   the v1 columns, with `versions` renamed `schema_versions`,
   `earliest_version` and `pct_filers_reporting` added and `duplicated` dropped);
   `format = "v2"` adds `family_id`, `part_id` and `multi_value`.

```{r build}
cc <- build_concordance(format = "v2")
dim(cc)
cc[variable_name == "F9_01_REV_TOT_CY", .(xpath, rdb_table, rdb_relationship, form_type)]
```

## One database per form

ef2 builds two research databases from the e-file XML: one for 990 and 990-EZ
returns (with their schedules) and one for 990-PF returns. Each gets its own
view of the concordance:

```{r forms}
f990 <- concordance(form = "F990")      # 990 and 990-EZ returns
f990pf <- concordance(form = "F990PF")  # 990-PF returns
c(all = nrow(concordance()), F990 = nrow(f990), F990PF = nrow(f990pf))
```

The 990-PF view holds the PF xpaths, the return header (the same `F9_00_`
variables as the 990) and Schedule B, which foundations also file. Attachments
filed with every form map to PF variables there (`xpath_forms.csv`):

```{r per-form}
x <- "/Return/ReturnData/ReasonableCauseExplanation/ExplanationTxt"
rbind(f990[xpath == x, .(view = "F990", variable_name, rdb_table)],
      f990pf[xpath == x, .(view = "F990PF", variable_name, rdb_table)])
```

## Functions for parsers and data users

```{r api}
# the minimal mapping a parser needs
head(xpath_map(form = "F990"), 3)

# one-to-many tables
head(table_names("MANY"))

# clean xpaths from an XML document and look them up
lookup_xpath("/irs:Return/irs:ReturnData/IRS990/Form990PartVIISectionAGrp[2]/PersonNm")

# one row per variable, in form order
dd <- data_dictionary("F990")
dd[part_id == "F9-P01", .(variable_name, label, data_type_simple)][1:5]
```

`cc_forms()`, `cc_parts()`, `cc_tables()`, `cc_variables()` and
`cc_xpaths()` return the component tables themselves.

## How a parser uses the concordance

For each return, a parser flattens the XML into (xpath, value) pairs, removes
the `[n]` indexes and namespace prefixes (`normalize_xpath()`), joins the
xpaths to its database's concordance view, and writes the values into tables:

* In a **ONE** table each variable is one column of the filing's row.
* In a **MANY** table each instance of the repeating group is a row.
* A **`multi_value`** variable can repeat inside a one-row table (for example
  the states where the return is filed); its values are joined into one cell.
* Xpaths ending in **`/@name`** are XML attributes of the element before
  them (for example the 501(c) subsection of `Organization501c`).
* A **`production_rule`** on an xpath describes any other special handling.

## Generated files

`data-raw/build-concordance.R` writes the files kept at the top of the
repository and stops if anything is wrong:

1. `check_changelog()`: every difference from v1 is in the change log, and
   replaying the log on v1 reproduces the current tables.
2. `validation_summary()`: every structural rule passes.
3. `concordance.csv` and `concordance.xlsx`: all xpaths, v1 layout.
4. `concordance-990pf.csv`: the 990-PF view, v1 layout.
5. `inst/extdata/changelog/v1_to_v2_crosswalk.csv`: `v1_crosswalk()`.

```{r write, eval = FALSE}
source("data-raw/build-concordance.R")
# or, for one file
write_concordance(build_concordance(format = "v1", form = "F990PF"), "concordance-990pf.csv")
```
