---
title: "Looking up tables and variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Looking up tables and variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
library(concordance990)
library(data.table)
```

The research databases built from the concordance have hundreds of tables and
thousands of variables. `dd()` answers the question you have while working
with them: *what is this column?* Give it a table name or a variable name and
it returns the data dictionary entry: label, description, the line of the
form it comes from (location code), which returns report it (scope), the data
type, and how many xpaths are pooled into it.

`dd()` reads the same dictionary as `data_dictionary()`, which returns the
whole dictionary of one database. The [Form 990](data-dictionary-990.html)
and [990-PF](data-dictionary-990pf.html) dictionary pages are built
from it too, so the three always agree.

## Look up a variable

A variable prints as a record card:

```{r}
print(dd("F9_01_REV_TOT_CY"))
```

* **Location** is the form, return type, part and line the value comes from
  (`F990-PC-PART-01-LINE-12-CY` is Form 990, Part I, line 12, current year).
* **Scope** says which returns have the variable: `PC` full 990 only, `EZ`
  990-EZ only, `PZ` both, `HD` the return header shared by all returns, `PF`
  the 990-PF. A `PC` variable is missing for 990-EZ filers.
* **Database** is the research database the variable belongs to (`F990` or
  `F990PF`). Header and Schedule B variables belong to both.
* **Forms** are the return types whose xpaths feed the variable, and
  **Xpaths** counts them.

Names are matched exactly, then ignoring case, so `dd("f9_01_rev_tot_cy")`
works too. Several variables print as several cards:

```{r}
print(dd(c("F9_01_REV_TOT_CY", "F9_01_EXP_TOT_CY"), fields = c("label", "location_code_family")))
```

`fields` picks the columns to return; `fields = "all"` returns every column
of the dictionary, and `names(dd("F9_00_ORG_EIN"))` lists them.

### Where does the value come from? Ask for the xpaths

The xpaths are not reported by default. Set `xpaths = TRUE` to list every
xpath pooled into the variable, with the schema years it appears in and the
percent of filers of that return type who report it:

```{r}
print(dd("F9_01_REV_TOT_CY", xpaths = TRUE))
```

The IRS renamed most elements in 2013, so a variable usually pools a
2009-2012 xpath and a 2013-2024 xpath, for each return type. Xpaths with
no years (`-`) are in the IRS schemas but have not been seen in filings.

### Questions about the raw data? Open the validation page

Every variable has a validation page built from the e-filed returns
(TY2009-2024): the value checks and flags with the reviewer's decisions, the
xpaths and the years each was used, filings and fill rate by tax year, a
summary of the values, and example filings with links to the XML.
`report = TRUE` opens it in the browser:

```{r, eval = FALSE}
dd("F9_01_REV_TOT_CY", report = TRUE)
```

`variable_page_url()` gives the address without opening it:

```{r}
variable_page_url("F9_01_REV_TOT_CY")
```

### Families of alternate versions

A few lines of the form are asked differently on the 990 and the 990-EZ, so
they are separate variables linked in a **family**. The card shows the family
and its definition:

```{r}
print(dd("F9_04_SCHED_B_REQ_X"))
```

## Look up a table

A table prints as a listing of its variables, in the order they appear on the
form, under a header with the part, the database and whether the table has
one row per filing (`ONE`) or one row per repeated item (`MANY`):

```{r}
print(dd("F9-P07-T02-CONTRACTORS"))
```

The listing fits the console. When the console is too narrow for the label to
share a line with the other columns, the label moves under each row, as above.
In a wider console it is a column:

```{r}
print(dd("F9-P07-T02-CONTRACTORS"), n = 4, width = 140)
```

Long tables show the first 50 variables; `n` changes that. `description =
TRUE` prints each variable's description under it:

```{r}
print(dd("F9-P01-T00-SUMMARY"), n = 3, description = TRUE)
```

With `xpaths = TRUE` a table lookup lists the xpaths of the variables shown.

A multi-value variable (`multi`) is a field that repeats inside a one-row
table, such as the states where the return is filed; parsers join the values
into one cell. The column only appears when the table has one.

## When a name is not found

A misspelled name is an error that suggests the closest names:

```{r, error = TRUE}
dd("F9_00_EIN")
dd("F9-P01-T00-SUMARY")
```

`form` limits the search to one database:

```{r, error = TRUE}
dd("PF-P01-T00-REVENUE-EXPENSE", form = "F990")
```

One call looks up either tables or variables. To see a variable in its table,
look up the variable, then its table:

```{r}
v <- dd("F9_06_DISCLOSURE_STATES_FILED")
v$rdb_table
```

## The result is a data.table

What prints is formatting only. `dd()` returns a data.table, one row per
variable, so you can filter it, join it to your data or write it out:

```{r}
x <- dd("F9-P01-T00-SUMMARY")
x[variable_scope == "PZ", .(variable_name, label)][1:5]
```

Once you select columns the result prints as a plain data.table, as above.
The xpaths are in an attribute:

```{r}
xp <- attr(dd("F9_01_REV_TOT_CY", xpaths = TRUE), "xpaths")
xp[, .(form_type, earliest_version, latest_version, pct_filers_reporting)]
```

To find variables by keyword, search the whole dictionary with
`data_dictionary()` and pass the names to `dd()`:

```{r}
d <- data_dictionary("F990")
hits <- d[grepl("volunteer", label, ignore.case = TRUE), variable_name]
print(dd(hits, fields = c("label", "rdb_table")))
```

## In R Markdown

Printed automatically in an R Markdown document (without `print()`), a `dd()`
result renders as markdown tables instead of console text, so the same call
documents variables in a report:

```{r}
dd("F9_01_REV_TOT_CY", xpaths = TRUE)
```

```{r}
dd("F9-P07-T02-CONTRACTORS", fields = c("label", "data_type_simple"))
```
