---
title: "From the v1 concordance to v2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From the v1 concordance to v2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(concordance990)
library(data.table)
n <- function(x) format(x, big.mark = ",")
v1 <- read_cc_csv(system.file("extdata", "v1", "concordance-v1.csv", package = "concordance990"))
s <- read_src()
ch <- read_changelog()
cw <- read_cc_csv(system.file("extdata", "changelog", "v1_to_v2_crosswalk.csv", package = "concordance990"))
```

## Why a new version

The version 1 Master Concordance File (`concordance.csv` in the
`irs-efile-master-concordance-file` repository) mapped
`r n(nrow(v1))` xpaths to `r n(uniqueN(v1$variable_name))` variables in
`r uniqueN(v1$rdb_table)` tables. It has been the backbone of the e-file
research databases, but its single flat file made it hard to maintain and to
check:

* **Facts were repeated and disagreed.** Every row carried its variable's
  description, table, label and scope. For 829 variables the xpaths gave
  different descriptions; 15 variables were assigned to more than one table.
* **Loose vocabularies.** Missing values were sometimes empty and sometimes
  the text `NA` (23,528 cells); a cardinality was `"ONE "` with a trailing
  space, so parsers skipped three Schedule H tables; identifiers such as EINs
  and ZIP codes were typed numeric.
* **No evidence loop.** Nothing tested a mapping against real filings. When
  the xpaths of every variable were compared with 5.98 million 990 and
  990-EZ filings (TY2009-2024), the checks found opposite meanings pooled
  into one variable (Schedule B *required* with Schedule B *not required*),
  revenue lines in expense variables, repeating groups in one-row tables that
  silently dropped values, and hundreds of observed fields with no mapping.
* **Every downstream package patched its own copy.** ef2, panel990, fiscal
  and efile-rdb-tables each re-derived table headers, types and fixes.
* **The 990-PF lived elsewhere**, in a separate concordance with its own
  naming.

## Design decisions

| | Decision |
|---|---|
| D1 | 990 and 990-EZ xpaths are pooled under one variable when they measure the same thing; non-equivalent lines become sibling variables. |
| D2 | The 990-PF is in scope, merged as form `F990PF`. |
| D3 | The component CSVs in GitHub are the source of truth; a v1-shaped `concordance.csv` (and `.xlsx`) is always generated. |
| D4 | The concordance ships as an R package that partner packages import. |
| D5 | Evidence comes from the e-file DuckDB builds, joined to the current concordance by cleaned xpath. |
| D6 | A family is named after its anchor variable. |
| D7 | Who files a part is declared per part and checked against filings. |
| D8 | Every change from v1 is logged line by line, so data built with v1 stays reconcilable. |

## How the migration was done

1. **Baseline.** `split_v1()` decomposed the frozen v1 file into the
   component tables without changing a value. Where v1 disagreed with itself,
   the most common value went to the variable and the rest were kept as
   xpath overrides. The v1 layout rebuilt from this baseline is byte-identical
   to v1 (a test checks it), and it is tagged `v2.0.0-baseline` on GitHub.
2. **Evidence.** Filing evidence was built from the DuckDB builds: for every
   xpath and year, how many filings use it, how often it repeats, and a
   profile of its values. Checks turn this into an issues list
   (`reports/issues.html`: `r n(29937)` cases).
3. **Fixes, each logged.** Every fix is a script in `data-raw/fixes/` that
   edits the component tables and appends one change-log row per changed
   cell, with a reason and the evidence behind it. Reviewed cases that are
   correct as they are go to the validation log instead.
4. **990-PF merge.** The PF concordance (`F990-PF-FULL.xlsx`) was split the
   same way and merged as form `F990PF`, then checked against the TY2023 PF
   filings and the 2023 schemas.

`reports/issues-resolved.html` lists every case of the issues list with what
was done and where it landed.

## What changed

```{r changes}
ch[, .(rows = .N), by = .(level, change_type)][order(-rows)][1:12]
ch[, .N, by = affects_data]
```

`affects_data = TRUE` marks changes that change data built with v1 (remaps,
splits, retypes, table moves, renames); `FALSE` marks metadata such as labels,
whitespace and the `NA` clean-up.

Today the concordance has `r n(nrow(s$xpaths))` xpaths,
`r n(nrow(s$variables))` variables and `r n(nrow(s$tables))` tables, including
the 990-PF (`r n(sum(s$xpaths$form == "F990PF"))` xpaths).

## Reconciling data built with v1

The crosswalk (`inst/extdata/changelog/v1_to_v2_crosswalk.csv`, rebuilt with
`v1_crosswalk()`) has one row per xpath in v1 or now, with the v1 and current
variable and table:

```{r crosswalk}
cw[, .N, by = status][order(-N)]
cw[status == "remapped", .(xpath, v1_variable_name, variable_name)][1:5]
```

To re-map a dataset built with v1, join its columns through the crosswalk:
`remapped` and `moved_table` rows move values, `added` rows are new
coverage, and `metadata_changed` rows need no data change. The change-log rows
with `affects_data = TRUE` explain each move.

## Changes that affect downstream code

* **Renamed variables**: names with trailing spaces or lower case,
  `SH_01_CHNA_DESC_RESOURCES_X` to `SH_05_...`, `SB_01_CONTRIBUTOR_TYPE` to
  `SB_01_CONTRIBUTOR_NUM`, and the 990-PF Part IX-A slots
  (`PF_09_CHARIT_ACT_DESC_1` ... `_4`).
* **Renamed table**: `SO-T99-SUPPLEMENTAL-INFO` is now
  `SO-P00-T99-SUPPLEMENTAL-INFO`.
* **New and re-shaped tables**: supplemental-information tables and
  per-facility tables are MANY; new tables hold Schedule H facility policies,
  Schedule A hospitals and agricultural research colleges, group-return
  subordinates, Schedule C affiliated groups and the 990-PF Schedule B.
* **The 990-PF is included.** `concordance.csv` contains the 990-PF rows;
  parsers select a database with `concordance(form = "F990")` or
  `concordance(form = "F990PF")`.
* **New conventions for parsers**: `multi_value` fields join repeated values,
  and xpaths ending in `/@name` are XML attributes.
