---
title: "The component tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The component tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(concordance990)
library(data.table)
s <- read_src()
n <- function(x) format(x, big.mark = ",")
```

The version 1 Master Concordance was one flat file with one row per xpath.
Every row repeated the attributes of its variable, table, part and form, so
the same fact was stored many times and could disagree with itself. Version 2
decomposes that file into small **component tables**, one per level of the
hierarchy. Each fact is stored once, and the familiar one-row-per-xpath file
is generated from the components (see `vignette("building")`).

The component tables are CSV files in `inst/extdata/src/`. They are the
source of truth: edit these, never the generated `concordance.csv`.

## The hierarchy

```
form        F990, F990PF, SCHED-A ... SCHED-R
 └ part     F9-P07  (Form 990 Part VII)
    └ table F9-P07-T01-COMPENSATION       one row per filing (ONE) or many (MANY)
       └ variable  F9_07_COMP_DTK_NAME_PERS
          └ xpath  /Return/ReturnData/IRS990/Form990PartVIISectionAGrp/PersonNm
                   /Return/ReturnData/IRS990/Form990PartVIISectionA/NamePerson   (2009-2012)
```

A **variable** is one research concept, a column in a table. It pools every
xpath that records that concept: the 990 and 990-EZ versions of a line and
the element names of different schema years (the IRS renamed most elements in
2013). An **xpath** belongs to exactly one variable; a variable belongs to
exactly one table.

## The tables

```{r counts, echo = FALSE}
d <- as.data.table(list(
  file = paste0(names(s), ".csv"),
  key = c("form_id", "part_id", "table_id", "variable_name", "xpath", "xpath + field", "family_id", "xpath + form_id"),
  rows = sapply(s, nrow),
  holds = c("Forms and schedules, with the XML root elements of each",
            "Parts of each form (Part I, Part II, ...), with titles",
            "Relational tables: part, cardinality (ONE / MANY)",
            "Variables: table, label, description, data type, scope, location code, list-valued flag",
            "Xpaths: variable, form line, schema versions and other xpath-level metadata",
            "Where one xpath of a variable differs from the variable value (kept from v1)",
            "Groups of alternate versions of one form line (e.g. _V2 variables)",
            "Form-specific mappings for returns parsed as a separate database")))
knitr::kable(d[, .(File = file, Key = key, Rows = n(rows), Holds = holds)])
```

### forms.csv

One row per form: the Form 990 (with the 990-EZ, whose lines are mapped into
the matching 990 parts), the 990-PF and Schedules A to R. `xml_roots` lists
the root elements of the form's xpaths (for example `IRS990`, `IRS990EZ`,
`ReturnHeader` and attachments such as `CompensationExplanation` for the
990). The validation rule R08 checks every xpath against these roots.

### parts.csv and tables.csv

A part is a numbered part of a form (`F9-P08`: Form 990 Part VIII, Statement
of Revenue). Parts `P00` hold the heading of a form; parts `P99` hold lines
from earlier versions of a schedule, or the auxiliary schedules of the 990-PF.

A table (`rdb_table`) is the unit a parser writes:

* **ONE** tables have one row per filing (`F9-P08-T00-REVENUE`).
* **MANY** tables have one row per repeating group in a filing: a person in
  Part VII, a grant recipient in Schedule I, a hospital facility in Schedule
  H (`F9-P07-T01-COMPENSATION`). By convention `T00` tables are ONE and
  `T01`+ tables MANY; `T99` tables hold supplemental explanations.

```{r cardinality}
cc_tables()[, .N, by = .(form = form_id, cardinality)][order(form, cardinality)][1:8]
```

### variables.csv

```{r variables}
cc_variables()[variable_name %in% c("F9_08_REV_TOT_CY", "F9_06_DISCLOSURE_STATES_FILED"),
               .(variable_name, table_id, label, data_type_simple, variable_scope, multi_value)]
```

* **Names** follow `FF_PP_NAME`: a two-character form code (`F9` Form
  990/990-EZ, `SA`-`SR` Schedules A to R, `PF` Form 990-PF), the two-digit
  part, and a descriptive name, at most 32 characters. 990-PF auxiliary
  schedules use `PF_AXnn_` for schedule `nn` (table `PF-P99-Tnn-...`).
* **`variable_scope`**: `HD` header (shared by all returns), `PC` full 990
  only, `EZ` 990-EZ only, `PZ` both, `PF` 990-PF.
* **`data_type_simple`**: `numeric`, `text`, `checkbox` or `date`.
  Identifiers (EIN, ZIP, phone) are text so leading zeros survive.
* **`multi_value`**: `TRUE` for fields whose element repeats inside a
  one-row table (states where the return is filed, foreign countries);
  parsers join the values into one cell.
* **`family_id`**: the variable itself, or the anchor variable of a family of
  alternate versions of one line (`families.csv`).

### xpaths.csv

One row per xpath, the primary key of the concordance. Besides the variable it
records the form, form type (`PC`, `EZ`, `SZ` schedule, `PF`), form part and
line, the IRS data type, the schema versions it appears in, and a free-text
`production_rule` where a value needs special handling (for example the TY2009
501(c)(3) flag, which is an attribute of `Organization501c`). Xpaths ending in
`/@name` are XML attributes.

`v1_order` keeps the row order of the v1 file, so the generated file stays in
the familiar order.

### xpath_overrides.csv

When v1 was split, each variable-level column got the most common value
across the variable's xpaths. Where an xpath disagreed (a different
description, a different table), the difference was kept here instead of
being lost. The build applies overrides after the join. Resolving a conflict
means deleting its override row, with a change-log entry.

```{r overrides}
cc_xpaths()[, .N]; read_src()$xpath_overrides[, .N, by = field][order(-N)]
```

### xpath_forms.csv

Some attachments are filed with every form (compensation explanations,
reasonable-cause statements). In 990 returns they belong to 990 variables; in
990-PF returns, which are parsed as a separate database, to PF variables. An
xpath has one primary mapping in `xpaths.csv`; `xpath_forms.csv` gives the
mapping for another form, applied by `concordance(form = "F990PF")`.

```{r xpath-forms}
read_src()$xpath_forms[1:3]
```

## Around the component tables

| File | Role |
|---|---|
| `inst/extdata/changelog/changes.csv` | Every cell changed since v1, with reason and evidence (`vignette("migration")`) |
| `inst/extdata/changelog/v1_to_v2_crosswalk.csv` | Each v1 xpath with its v1 and current variable and table |
| `inst/extdata/validation/validation_log.csv` | Reviewer decisions on flagged cases (`vignette("testing-updates")`) |
| `inst/extdata/validation/rule_ceilings.csv` | Ceilings for structural rules still being fixed (none at present) |
| `inst/extdata/v1/concordance-v1.csv` | The frozen v1 file, byte for byte |
| `concordance.csv`, `concordance.xlsx` | Generated: all xpaths in the v1 column layout |
| `concordance-990pf.csv` | Generated: the 990-PF view |

## Reading the tables

```{r join}
x <- merge(cc_xpaths()[, .(xpath, variable_name)],
           cc_variables()[, .(variable_name, table_id, label)], by = "variable_name")
x <- merge(x, cc_tables()[, .(table_id, cardinality, part_id)], by = "table_id")
x <- merge(x, cc_parts()[, .(part_id, title)], by = "part_id")
x[variable_name == "F9_01_REV_TOT_CY", .(xpath, table_id, cardinality, title)]
```

`concordance()` does this join (and applies the overrides) for you.
