---
title: "Validation, tests and making updates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation, tests and making updates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(concordance990)
library(data.table)
```

The concordance is checked at three levels. **Structural rules** test the
component tables themselves and run on every build and in CI. **Evidence
flags** and **value checks** compare the mapping with real filings. Every
change is logged, so any version can be traced back to v1.

## 1. Structural rules

`validate_concordance()` applies twelve rules to the component tables; all are
enforced, so a violation stops the build and fails the tests.

```{r rules}
knitr::kable(validation_summary()[, .(rule, status, violations, description)])
```

Two rules accept documented exceptions: R04/R06 allow the 990-PF auxiliary
schedule names `PF_AXnn_` in `PF-P99` tables, and R07 skips variables whose
use of several tables is accepted in the validation log (the Form 990 Part III
program tables share variable names so they can be stacked).

When a rule still has known violations it can be run as a **ratchet**: its
ceiling in `inst/extdata/validation/rule_ceilings.csv` may go down but never
up, and at zero the rule becomes enforced.

## 2. Evidence from filings

`build_evidence()` scans the e-file DuckDB builds (one per tax year) and
writes compact tables to `evidence/`: for every xpath, year, schema version
and return type, the number of filings, how often it repeats within a filing,
the repeating element it sits in, and a profile of its values. From these,
`flag_xpaths()` raises flags and `run_value_checks()` runs the checks shown in
the per-variable validation reports.

```{r catalog}
knitr::kable(issue_catalog[, .(check, severity, name)])
```

A few examples of what they catch:

* **COLLISION**: two xpaths of one variable filled in the same row (a
  repeating-group instance, or the filing), so a parser would drop one. It
  found book value pooled with fair market value and a revenue line pooled
  with an expense line.
* **POLARITY**: `...NotRequired` pooled with `...Required`.
* **ONE_REPEATS**: an xpath that repeats inside a one-row table, whose extra
  values would be lost.
* **UNMAPPED** / **GAP**: fields that appear in filings but not in the
  concordance, and years where a variable has no data because its successor
  element is not mapped.

`collect_issues()` gathers every failing case with the file, row and column
to edit, and `write_issues_report()` writes `reports/issues.html`.

## 3. The validation log

Not every flag is an error: a count whose median moves with the filer
population, an address field whose foreign values do not look like ZIP codes.
Reviewed cases are recorded in `inst/extdata/validation/validation_log.csv`,
one row per decision:

```{r log}
vl <- read_validation_log()
vl[, .N, by = status]
vl[check == "POLARITY"][1, .(check, variable_name, status, note)]
```

`accepted` cases (correct as they are, or a false positive of the check)
drop out of later issues lists; `needs_input` and `deferred` cases stay.

## 4. Tests and CI

`devtools::test()` runs the testthat suite in `tests/testthat/`:

| File | Tests |
|---|---|
| `test-structure.R` | keys are unique, references resolve, controlled values |
| `test-validate.R` | every rule passes; the rules catch deliberate errors |
| `test-changelog.R` | the change log replays exactly; add, edit and remove round-trip; stale old values are rejected; the crosswalk covers every xpath |
| `test-roundtrip.R` | the baseline split reproduces v1 byte for byte |
| `test-api.R` | the consumer functions, the per-form views and the data dictionary |

GitHub Actions runs `R CMD check` (which runs the tests and builds these
vignettes) on every push and pull request (`.github/workflows/R-CMD-check.yaml`).

## 5. Making an update

Edit the component tables, never the generated files. Each fix is logged, one
change-log row per changed cell (`inst/extdata/changelog/changes.csv`), with a
reason and the evidence behind it, stored once per set of changes in
`change_sets.csv` (`read_changelog()` joins them). Derived columns
(`derived_columns`: the xpath version fields and `pct_filers_reporting`) are
recomputed from the filing evidence and not logged cell by cell. Replaying the log on the frozen v1 must
reproduce the current tables exactly; the build stops if it does not.

### With a fix script (recommended)

The fixes so far are scripts in `data-raw/fixes/`. Each step edits the
tables and logs its own changes:

```{r fix-script, eval = FALSE}
source("data-raw/fixes/_helpers.R")   # fix_step(), rename_variable(), remap_xpaths(), ...

fix_step(function(s) {
  remap_xpaths(s, "/Return/ReturnData/IRS990ScheduleD/RevenueNotRptdOnFinStmt",
               "SD_11_RECO_REV_ADD_L4AB")
}, reason = "Remap the revenue reconciliation total from the expense variable to the revenue variable.",
   evidence = "POLARITY; COLLISION; reports/SD_12_RECO_EXP_ADD_L4AB.html",
   type = "remap", affects = TRUE)
```

### By hand

```{r by-hand, eval = FALSE}
# 1. edit inst/extdata/src/*.csv
# 2. draft change-log rows for the edits, then fill in reason and evidence
devtools::load_all()
draft_changes(author = "Your Name", reason = "...", evidence = "...", write = TRUE)
```

### Then

```{r then, eval = FALSE}
source("data-raw/build-concordance.R")   # checks the log and the rules, writes the files
devtools::test()
```

Record reviewed flags in the validation log, commit the component tables, the
change log and the generated files together, and open a pull request.

### Change types

```{r types}
change_types
```

Mark `affects_data = TRUE` when data built with v1 changes (a remap, split,
retype, table move or rename) and `FALSE` for metadata (labels, whitespace).
