---
title: "Data dictionary: Form 990 and 990-EZ"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("_dictionary.R")
dd <- data_dictionary("F990")
```

This dictionary documents every variable of the **990 / 990-EZ research
database**: the Form 990, the 990-EZ and Schedules A to R. It is generated
from the concordance with `data_dictionary("F990")`, so it always matches the
package. The 990-PF has [its own dictionary](data-dictionary-990pf.html); the
[outline](form-outline.html) lists every part and table on one page.

<a class='btn btn-outline-primary btn-sm' href='../print/data-dictionary-990.pdf'>Printable version (PDF)</a> &nbsp;one table on landscape letter pages, for printing or offline reading.

## Reading the dictionary

**Validation pages.** Click a variable name to open its [validation page](../variables/index.html): the xpaths pooled into it and the years each was used, coverage and fill rate by tax year, value checks and flags, and example filings with links to the XML. From R, `dd("<name>", report = TRUE)` opens the same page.

**Variables and fields.** A variable is not a field on one form. It pools
every xpath that records the same thing: the 990 and 990-EZ versions of a line
(for example total revenue), and the element names of different schema years
(the IRS renamed most elements in 2013). A
variable that exists only on the full 990 (the number of volunteers, say) is
missing for 990-EZ filers.

**Names** follow `FF_PP_NAME`:

* `FF` is the form: `F9` for the Form 990 and 990-EZ (and the return header),
  `SA` to `SR` for Schedules A to R.
* `PP` is the part of the form (`00` for the heading, `01` for Part I, ...;
  `99` for lines from earlier versions of a schedule).
* `NAME` describes the field; `_X` marks a checkbox, `_CY`/`_PY` current and
  prior year, `_BOY`/`_EOY` beginning and end of year.

**Scope** says which returns report the variable: *header* (all returns,
including the 990-PF), *990* (full form only), *990-EZ* (EZ only) or *990 +
990-EZ*. 990-EZ lines are mapped into the matching part of the 990.

**Tables.** Each variable is a column of one table. *ONE* tables have one row
per filing; *MANY* tables have one row per repeated item (a person, a grant
recipient, a facility). The Part III program tables share variable names on
purpose, so the programs can be stacked.

**Type** is `numeric`, `text`, `checkbox` or `date`; identifiers such as EINs
and ZIP codes are text. A <span class='badge bg-secondary'>list</span> marks a
field whose repeated values are joined into one cell (for example the states
where the return is filed).

**Location** is the form, part and line code of the variable. Each table is a block headed by its name and cardinality: ONE (one row per filing) or MANY (one row per repeated item).

**% Reporting** is the percent of filers in the variable's scope whose return
reports it in the newest schema year (TY2024). A *990 + 990-EZ* variable shows
two values, 990/990-EZ (for example `16.9%/3.2%`). `<0.1%` is under one filer
in a thousand; blank means the variable is not in the current schema.

## Contents

```{r summary}
dictionary_summary("F990")
```

```{r css, results = "asis"}
cat(dictionary_css())
```

```{r sections, results = "asis"}
cat(dictionary_sections("F990"))
```

