---
title: "Data dictionary: Form 990-PF"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("_dictionary.R")
```

This dictionary documents every variable of the **990-PF research
database**: the Form 990-PF, the auxiliary schedules and statements filed with
it, Schedule B and the return header shared with the 990. It is generated from
the concordance with `data_dictionary("F990PF")`, the variables of
`concordance(form = "F990PF")`. The Form 990 and 990-EZ have
[their own dictionary](data-dictionary-990.html).

<a class='btn btn-outline-primary btn-sm' href='../print/data-dictionary-990pf.pdf'>Printable version (PDF)</a> &nbsp;one table on landscape letter pages, for printing or offline reading.

## Reading the dictionary

**Validation pages.** Click a variable name to open its [validation page](../variables/index.html): the xpaths pooled into it and the years each was used, coverage and fill rate by tax year, value checks and flags, and example filings with links to the XML. From R, `dd("<name>", report = TRUE)` opens the same page.

**Names** follow `PF_PP_NAME`, where `PP` is the part of the 990-PF, numbered
as on the forms before the 2023 redesign (the part titles give the 2023
number). Two other prefixes appear:

* `F9_00_` for the **return header**, shared with the 990 and 990-EZ: the
  same variables in both databases.
* `PF_AXnn_` for **auxiliary schedule** `nn`, the statements a foundation
  attaches (depreciation, other income, grants approved for future payment,
  compensation explanations, and others), in tables `PF-P99-Tnn-...`.
* `SB_` for **Schedule B**, which foundations file too. Unlike the 990,
  990-PF Schedule B contributor names and addresses are public.

**Attachments filed with every form** (compensation explanations,
reasonable-cause statements) map to their `PF_AX` variables in this database
and to 990 variables in the 990 database.

**Tables.** *ONE* tables have one row per filing; *MANY* tables have one row
per repeated item (an officer, a grant paid, an asset sold). Fixed line slots
are columns of a one-row table, for example the four direct charitable
activities of Part IX-A (`PF_09_CHARIT_ACT_DESC_1` to `_4`).

**Type** is `numeric`, `text`, `checkbox` or `date`. A
<span class='badge bg-secondary'>list</span> marks a field whose repeated
values are joined into one cell (for example the manager names of Part XV).

**Evidence.** The 990-PF mapping was checked against one year of filings
(TY2023, 114,559 returns) and the 2023 schemas. Xpaths of earlier schema
versions are kept from the v1 PF concordance.

**% Reporting** is the percent of filers in the variable's scope whose return
reports it in the newest schema year (TY2024). `<0.1%` is under one filer in a
thousand; blank means the variable is not in the current schema.

## Contents

```{r summary}
dictionary_summary("F990PF")
```

```{r css, results = "asis"}
cat(dictionary_css())
```

```{r sections, results = "asis"}
cat(dictionary_sections("F990PF"))
```

