---
title: "Outline of forms, parts and schedules"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("_dictionary.R")
```

The IRS 990 returns are divided into **parts** that group fields by theme.
This page outlines every part of the Form 990 (with the 990-EZ), the
Schedules A to R and the Form 990-PF, and the concordance tables that hold
their fields. Table names follow `FF-PNN-TNN-NAME`: form, part, table number
(by convention `T00` tables have one row per filing, `T01` and higher one row
per repeated item, `T99` supplemental explanations) and a short name.

The part titles follow the IRS forms. Form 990-EZ lines are mapped into the
matching part of the Form 990. The 990-PF parts are numbered as on the forms
before the 2023 redesign, which moved most parts down by one number; the 2023
number is given in parentheses.

Variable counts link to the data dictionaries:
[Form 990 and 990-EZ](data-dictionary-990.html) and
[Form 990-PF](data-dictionary-990pf.html).

<a class='btn btn-outline-primary btn-sm' href='../print/form-outline.pdf'>Printable version (PDF)</a> &nbsp;one table on portrait letter pages, for printing or offline reading.

```{r css, results = "asis"}
# table names in the same bold monospace as the dictionary variable names
cat('<link rel="stylesheet" href="https://fonts.googleapis.com/css2?family=JetBrains+Mono:wght@700&display=swap">
<style>
main li code { font-family: "JetBrains Mono", ui-monospace, Consolas, monospace; font-weight: 700;
               font-size: .85em; background: transparent; color: #1b1f23; padding: 0; }
</style>')
```

```{r outline, results = "asis"}
tb <- cc_tables()
v990 <- data_dictionary("F990")[, .(n = uniqueN(variable_name)), by = rdb_table]
vpf <- data_dictionary("F990PF")[, .(n = uniqueN(variable_name)), by = rdb_table]
nv <- rbind(v990, vpf[!rdb_table %in% v990$rdb_table])
tb <- merge(tb, nv, by.x = "table_id", by.y = "rdb_table", all.x = TRUE)
tb[is.na(n), n := 0L]
pt <- cc_parts()
fm <- cc_forms()[order(as.integer(sort_order))]
# the 990-PF after the 990 and before the schedules it does not file
fm <- rbind(fm[form_id == "F990"], fm[form_id == "F990PF"], fm[!form_id %in% c("F990", "F990PF")])
for (f in fm$form_id) {
  cat(sprintf("\n## %s\n\n", fm[form_id == f, title]))
  cat(sprintf("*Filed with: %s.*\n\n", filed_with[[f]]))
  for (p in pt[form_id == f][order(part_id), part_id]) {
    cat(sprintf("* **%s**\n", pt[part_id == p, title]))
    for (t in tb[part_id == p][order(table_id), table_id]) {
      x <- tb[table_id == t]
      cat(sprintf("    * `%s` &nbsp;<span class='text-muted small'>%s, %s variables</span>\n",
                  t, x$cardinality, x$n))
    }
  }
}
```
